---
title: "What sample size to estimate infection prevelance, a simulation study"
author: "Roman Luštrik"
date: "2020-04-18"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = TRUE}
library(ggplot2)

source("simulation_functions.R", max.deparse.length = 1000, echo = TRUE)

options(scipen = 14)
```

In order to successfully determine level of prevalence of a disease on a sample of population, we need to gather a sufficiently large sample. Here, I define the power as percentage (out of $propN$) of cases when simulated value is included in 95% CI.

Algorithm:

1. sample $N$ individuals from a binomial distribution of size 1 and $P = myprob$
2. calculate confidence intervals and estimate of prevalence
3. repeat 1-2 $propN$ times -> $X$
4. calculate proportion of times and simulated $P$ is within the confidence intervals ($within_ci$) and its standard deviations for $X$

Parameters in:
 * simulated probability ($prob$)
 * simulated population size ($N$)
 * number of times simulation is run for each $N$ ($propN$)
 
 Parameters out:
 
 * parameter estimate $p$
 * lower confidence intervals of $p$
 * upper confidence interval of $p$
 * power (in %)

```{r}
probs1 <- c(0.0001, 0.0005, 0.001, 0.01, 0.1)
```

```{r, warning = FALSE}
sims1 <- runBatch(simseq = seq(from = 200, to = 800, by = 200),
                  myprob = probs1)
sims2 <- runBatch(simseq = seq(from = 1000, to = 10000, by = 1000),
                  myprob = probs1)
```

```{r}
sims_1_2 <- do.call(rbind, list(sims1, sims2))

ggplot(sims_1_2, aes(x = N, y = power, color = as.factor(prob))) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Set1", name = "simulated prevalence") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.2)) +
  geom_vline(xintercept = 3000, color = "grey40") +
  geom_line() +
  geom_point(size = 3)
```

Each point represents average of 100 ($propN$) simulations where simulated value has been covered by 95% confidence interval - power.

Notice how power is increasing slowly for simulated prevalence 0.0001 and doesn't reach satisfactory power even after sample size of 10000. For simulated prevalence 0.0005 and 0.001 there is a steep curve that reaches power of 0.8 around 3000 samples (vertical line). For simulated prevalences $\geq 0.01$, power is nearly constantly over the 0.9 line.